Hydrodynamics of a Semipermeable Inextensible Membrane Under Flow and Confinement

### 2021-02-15 vesicle_code/docs/permeableVes/semipermeable_vesicle.tex
edited up to start of section iv

\section{Formulation\label{sec:formulation}}
We consider a two-dimensional semipermeable vesicle (permeable only to
water) suspended in a viscous fluid (Figure~\ref{fig:sketch}), and we
let $\gamma$ denote the membrane boundary, of length $L$ and enclosing an area $A$.
%
%$\alpha=4\pi A/L^2$ be the vesicle's reduced area where $A$ is its area
%and $L$ is its length. 
%
To focus on the effect of water flow on vesicle
hydrodynamics in various conditions, we assume the fluid is the same
inside and outside the vesicle membrane. On the membrane the permeable
water flux gives rise to a difference between the fluid velocity
${\uu}({\xx}\in \gamma)$ and the membrane velocity
$\dot{\xx}$~\cite{yao-mor2017}:
\begin{align}
\label{eqn:water_flux}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\nn$ is the outward normal (Figure~\ref{fig:sketch}), $k_w$ is
the hydraulic conductivity (m$^2$s/kg), $R$ is the ideal gas constant,
$T$ is the temperature (K), $\triangle c$ is contrast in a solute
concentration (mol/volume), and $\ff_\mathrm{mem}$ is the membrane
stress that consists of both bending $ \ff_\mathrm{ben}= -k_b
\xx_{ssss}$ and tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$, where $s$
is the arclength, $\sigma = \Lambda - \frac{3}{2}\kappa^2$, $\Lambda$ is the
membrane tension, and $\kappa$ is the membrane curvature.

@@

For example, we use $10^{-5}\le\beta\le10^{-2}$
in our simulations of a semipermeable inextensible membrane in free space under various flowing conditions.
We found that 
the equilibrium shape of a semipermeable membrane in free space to be independent of $\beta$.
Within this range of $\beta$ we find that $\beta$ only affects the time it takes for the membrane to reach the equilibrium shape, which is determined by the background flow
(such as a quiescent flow, a linear shear flow and its shear rate, and a Poiseuille flow and its maximum velocity.)
From these results and our analysis of area dynamics, we suspect similar hydrodynamics of a semipermeable vesicle
(with $10^{-11}\le\beta\le 10^{-9}$). 
%
When we use $\beta\ge 0.1$, we find that the semipermeable membrane swells to maximize its water content, independent of the initial reduced 
area and the background flow.
%
%In the simulations we focus on intermediate value of $10^{-5}\le \beta\le 10^{-3}$ to investigate the hydrodynamics of a semipermeable inextensible membrane. 
%

@@

\subsection{Equilibrium and Areal Dynamics in a Quiescent Flow\label{subsec:area_dynamics}}
In the absence of an imposed flow, the steady equilibrium shape and a
differential equation for the area inside the semipermeable inextensible membrane can be derived
by following the analysis of Veerapaneni {\em et al.}~\cite{vee-raj-bir-pur2009} in a quiescent flow $({\bf u}_{\infty} = 0)$.
%
%by computing the steady state shape of a
%semipermeable vesicle in a quiescent flow. 
%
The steady state shape is a
fixed point of the Lagrangian
\begin{align}
  \mathcal{L} = \frac{1}{2}\int_{\gamma} \kappa^2 \, ds +
    \Lambda \left(\int_{\gamma} ds  - L \right),
\end{align}
where the Lagrange multiplier $\Lambda$ is the tension. Note the absence
of the pressure Lagrange multiplier since the total water content inside the vesicle membrane is not conserved.
Taking the variation of $\mathcal{L}$ with respect to $\gamma$ and
setting it to zero, the steady state shape satisfies $\kappa_{ss} +
\frac{1}{2}\kappa^3 - \Lambda \kappa = 0$, with initial conditions
$\kappa(0) = \kappa_0$ and $\kappa_s(0) = 0$.  Multiplying by $\kappa_s$
and integrating once, we have
\begin{align}
  \frac{\kappa_s^2}{2} + \frac{\kappa^4}{8} - 
    \frac{\Lambda}{2}\kappa^2 = \frac{\kappa_0^4}{8} - 
    \frac{\Lambda}{2}\kappa_0^2.
  \label{eqn:curveODE}
\end{align}
Assuming $\kappa_s \neq 0$, Equation~\eqref{eqn:curveODE} is separable,
and can be solved analytically in terms of the {\em EllipticF} function
using Mathematica. Unlike in the impermeable case, the solution of
Equation~\eqref{eqn:curveODE} is not periodic unless $\kappa(s) =
\kappa_0$. Therefore, the equilibrium shape of a semipermeable vesicle
in a quiescent flow is a circle with radius $L/2\pi$ and constant
tension $\Lambda = \kappa^2/2 = 2\pi^2/L^2.$ Note that in contrast to a
impermeable vesicle, the steady state tension is positive rather than
negative. Such equilibrium shape of a semipermeable vesicle results from
the balance between elastic stress and membrane tension. For an elastic
membrane with a large pore, the equilibrium membrane (hemispherical)
shape also results from such a balance~\cite{Ryham2018_JFM}.

We next compute the area dynamics of a semipermeable vesicle. Since the
area of the vesicle is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds,
  \label{eqn:area}
\end{align}
its derivative includes terms due to the time derivative of $\xx \cdot
\nn$ and a term due to interface stretching~\cite{lai-tse-hua2008}, but
this latter term is zero because of the local inextensibility condition.
Therefore,
\begin{align}
  \dot{A}(t) =
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
Applying integration by parts, the two integrals are identical and we combine them to obtain
\begin{align}
  \dot{A}(t) = \int_\gamma \left(\beta (\ff \cdot \nn)\nn 
    + \SS[\ff]\right) \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where we have used the incompressibility of the single-layer potential.
Substituting in $\ff = \ff_\mathrm{ben} + \ff_\mathrm{ten}$ and
repeatedly applying integration by parts, we obtain
\begin{align}
  \dot{A}(t) & = \beta \int_\gamma \left(
    \frac{\kappa^3}{2} - \kappa \Lambda \right) \, ds.
  \label{eqn:areaROC}
\end{align}
%\begin{align}
%  \dot{A}(t) &= \beta \int_\gamma \left( -\xx_{ssss} + 
%    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
%  &= -\beta \int_\gamma \left(-\xx_{sss} + \sigma \xx_s 
%    \right) \cdot \kappa \xx_s \, ds \\
%  &= -\beta \int_\gamma \left(\xx_{ss} \cdot 
%    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
%  &= -\beta \int_\gamma \left(\kappa (\xx_{ss} \cdot \xx_{ss}) + 
%    \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
%    \, ds \\
%  &  = -\beta \int_\gamma \left(\kappa^3 + \kappa \sigma \right) 
%    \, ds = \beta \int_\gamma \left(
%    \frac{\kappa^3}{2} - \kappa \Lambda \right) \, ds.
%  \label{eqn:areaROC}
%\end{align}
Note that the integrand in Equation~\eqref{eqn:areaROC} is zero when the membrane tension balances the elastic force
\begin{align}
\label{eqn:areaROC2}
\Lambda &= \frac{\kappa^2}{2}=\frac{\kappa_0^2}{2},
%  \;\mathrm{ at }\; \mathrm{equilibrium},
\end{align}
which is consistent with the equilibrium results in Equation~(\ref{eqn:curveODE}).

@@

\subsection{Discretization in Time}
We apply a second-order spectral deferred correction (SDC) method that was
thoroughly tested in previous work~\cite{qua-bir2016}. To control error
over long time horizons, we use the error in length to adaptively adjust
the time step size. SDC requires two steps of a first-order time
stepping method. To allow for large time step sizes, terms involving
high-order derivatives are discretized semi-implicitly. We start by
defining
\begin{align}
  \BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  \TT\sigma = (\sigma \xx_s)_s, \quad
  \DD\ff = \xx_s \cdot \ff_s, \quad
  P\ff = (\ff \cdot \nn) \nn,
  \label{eqn:operators}
\end{align}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} can be written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(\BB\xx + \TT\sigma)
  + \SS (\BB\xx + \TT\sigma), \quad \DD \dot{\xx} = 0.
\end{align}
We note that the operators defined in Equation~\eqref{eqn:operators}, as
well as the single-layer potential defined in Equation~\eqref{eqn:SLP}
all depend on the vesicle shape. A time stepping method that treats
these dependencies implicitly would result in a non-linear set of
equations that would be computationally expensive to solve. Instead, we
use a time stepping method that defines the operators at time $t^N$ and
discretizes the stiff terms due to high-order derivatives
implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. In particular, the
first-order semi-implicit method that is applied iteratively with SDC is
\begin{alignat}{3}  
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right) 
  + \SS^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right),  \\
  \DD^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.

>>>>

\section{Formulation\label{sec:formulation}}
We consider a two-dimensional semipermeable vesicle (permeable only to
water) suspended in a viscous fluid (Figure~\ref{fig:sketch}), and we
let $\gamma$ denote the membrane boundary, of length $L$ and enclosing a
region of area $A$.
%
%$\alpha=4\pi A/L^2$ be the vesicle's reduced area where $A$ is its area
%and $L$ is its length. 
%
To focus on the effect of water flow on vesicle
hydrodynamics in various conditions, we assume the fluid is the same
inside and outside the vesicle membrane. On the membrane the permeable
water flux gives rise to a difference between the fluid velocity
${\uu}({\xx}\in \gamma)$ and the membrane velocity
$\dot{\xx}$~\cite{yao-mor2017}:
\begin{align}
\label{eqn:water_flux}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\nn$ is the outward normal (Figure~\ref{fig:sketch}), $k_w$ is
the hydraulic conductivity (m$^2$s/kg), $R$ is the ideal gas constant,
$T$ is the temperature (K), $\triangle c$ is contrast in a solute
concentration (mol/volume), and $\ff_\mathrm{mem}$ is the membrane
stress that consists of both bending $ \ff_\mathrm{ben}= -k_b
\xx_{ssss}$ and tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$, where $s$
is the arclength, $\sigma = \Lambda - \frac{3}{2}\kappa^2$, $\Lambda$ is the
membrane tension, and $\kappa$ is the membrane curvature.

@@

For example, we use $10^{-5}\le\beta\le10^{-2}$ in our simulations of a
semipermeable inextensible membrane in free space under various flowing
conditions. We found that the equilibrium shape of a semipermeable
membrane in free space to be independent of $\beta$. Within this range
of $\beta$ we find that $\beta$ only affects the time it takes for the
membrane to reach the equilibrium shape, which is determined by the
background flow (such as a quiescent flow, a linear shear flow and its
shear rate, and a Poiseuille flow and its maximum velocity). From these
results and our analysis of area dynamics, we suspect similar
hydrodynamics of a semipermeable vesicle (with $10^{-11}\le\beta\le
10^{-9}$). 
%
When we use $\beta\ge 0.1$, we find that the semipermeable membrane swells to maximize its water content, independent of the initial reduced 
area and the background flow.
%
%In the simulations we focus on intermediate value of $10^{-5}\le \beta\le 10^{-3}$ to investigate the hydrodynamics of a semipermeable inextensible membrane. 
%

@@

\subsection{Equilibrium and Areal Dynamics in a Quiescent Flow\label{subsec:area_dynamics}}
In the absence of an imposed flow, the steady equilibrium shape and a
differential equation for the area inside the semipermeable inextensible membrane can be derived
by following the analysis of Veerapaneni {\em et al.}~\cite{vee-raj-bir-pur2009} in a quiescent flow $({\bf u}_{\infty} = 0)$.
%
%by computing the steady state shape of a
%semipermeable vesicle in a quiescent flow. 
%
The steady state shape is a
fixed point of the Lagrangian
\begin{align}
  \mathcal{L} = \frac{1}{2}\int_{\gamma} \kappa^2 \, ds +
    \Lambda \left(\int_{\gamma} ds  - L \right),
\end{align}
where the Lagrange multiplier $\Lambda$ is the tension. Note the absence
of the pressure Lagrange multiplier since the total water content inside the vesicle membrane is not conserved.
Taking the variation of $\mathcal{L}$ with respect to $\gamma$ and
setting it to zero, the steady state shape satisfies $\kappa_{ss} +
\frac{1}{2}\kappa^3 - \Lambda \kappa = 0$, with initial conditions
$\kappa(0) = \kappa_0$ and $\kappa_s(0) = 0$.  Multiplying by $\kappa_s$
and integrating once, we have
\begin{align}
  \frac{\kappa_s^2}{2} + \frac{\kappa^4}{8} - 
    \frac{\Lambda}{2}\kappa^2 = \frac{\kappa_0^4}{8} - 
    \frac{\Lambda}{2}\kappa_0^2.
  \label{eqn:curveODE}
\end{align}
Assuming $\kappa_s \neq 0$, Equation~\eqref{eqn:curveODE} is separable,
and can be solved analytically in terms of the {\em EllipticF} function
using Mathematica. Unlike in the impermeable case, the solution of
Equation~\eqref{eqn:curveODE} is not periodic unless $\kappa(s) =
\kappa_0$. Therefore, the equilibrium shape of a semipermeable vesicle
in a quiescent flow is a circle with radius $L/2\pi$ and constant
tension $\Lambda = \kappa^2/2 = 2\pi^2/L^2.$ Note that in contrast to a
impermeable vesicle, the steady state tension is positive rather than
negative. The equilibrium shape of a semipermeable vesicle results from
the balance between elastic stress and membrane tension. For an elastic
membrane with a large pore, the equilibrium membrane (hemispherical)
shape also results from such a balance~\cite{Ryham2018_JFM}.

We next compute the area dynamics of a semipermeable vesicle. Since the
area of the vesicle is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds,
  \label{eqn:area}
\end{align}
its derivative includes terms due to the time derivative of $\xx \cdot
\nn$ and a term due to interface stretching~\cite{lai-tse-hua2008}, but
this latter term is zero because of the local inextensibility condition.
Therefore,
\begin{align}
  \dot{A}(t) =
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
Applying integration by parts, the two integrals are identical and we combine them to obtain
\begin{align}
  \dot{A}(t) = \int_\gamma \left(\beta (\ff \cdot \nn)\nn 
    + \SS[\ff]\right) \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where we have used the incompressibility of the single-layer potential.
Substituting in $\ff = \ff_\mathrm{ben} + \ff_\mathrm{ten}$ and
repeatedly applying integration by parts, we obtain
\begin{align}
  \dot{A}(t) & = \beta \int_\gamma \left(
    \frac{\kappa^3}{2} - \kappa \Lambda \right) \, ds.
  \label{eqn:areaROC}
\end{align}
%\begin{align}
%  \dot{A}(t) &= \beta \int_\gamma \left( -\xx_{ssss} + 
%    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
%  &= -\beta \int_\gamma \left(-\xx_{sss} + \sigma \xx_s 
%    \right) \cdot \kappa \xx_s \, ds \\
%  &= -\beta \int_\gamma \left(\xx_{ss} \cdot 
%    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
%  &= -\beta \int_\gamma \left(\kappa (\xx_{ss} \cdot \xx_{ss}) + 
%    \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
%    \, ds \\
%  &  = -\beta \int_\gamma \left(\kappa^3 + \kappa \sigma \right) 
%    \, ds = \beta \int_\gamma \left(
%    \frac{\kappa^3}{2} - \kappa \Lambda \right) \, ds.
%  \label{eqn:areaROC}
%\end{align}
Note that the integrand in Equation~\eqref{eqn:areaROC} is zero when the membrane tension balances the elastic force
\begin{align}
\label{eqn:areaROC2}
\Lambda &= \frac{\kappa^2}{2}=\frac{\kappa_0^2}{2},
%  \;\mathrm{ at }\; \mathrm{equilibrium},
\end{align}
which is consistent with the equilibrium result in
Equation~(\ref{eqn:curveODE}).

@@

\subsection{Discretization in Time}
We apply a second-order spectral deferred correction (SDC) method that
was thoroughly tested in previous work~\cite{qua-bir2016}. To control
error over long time horizons, we use the error in length to adaptively
adjust the time step size. Our SDC formulation requires two applications
of a first-order time stepping method. To allow for large time step
sizes, terms involving high-order derivatives are discretized
semi-implicitly. We start by defining
\begin{align}
  \BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  \TT\sigma = (\sigma \xx_s)_s, \quad
  \DD\ff = \xx_s \cdot \ff_s, \quad
  P\ff = (\ff \cdot \nn) \nn,
  \label{eqn:operators}
\end{align}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} can be written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(\BB\xx + \TT\sigma)
  + \SS (\BB\xx + \TT\sigma), \quad \DD \dot{\xx} = 0.
\end{align}
We note that the operators defined in Equation~\eqref{eqn:operators} and
the single-layer potential defined in Equation~\eqref{eqn:SLP} depend on
the vesicle shape. A time stepping method that treats these dependencies
implicitly would result in a non-linear set of equations that would be
computationally expensive to solve. Instead, we define the operators at
time $t^N$ and discretize the stiff terms due to high-order derivatives
implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. In particular, the
first-order semi-implicit method that is applied iteratively within SDC
is
\begin{alignat}{3}
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right) 
  + \SS^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right),  \\
  \DD^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.
